We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-21 paper/paper.tex
pdf

This started with the suggestion of a Magellanic Plane, a flattened
structure of satellite galaxies around the MW, first suggested by
\cite{1976RGOB..182..241K} and \cite{1976MNRAS.174..695L}.  
\cite{2005A&A...431..517K} then quantified that the
highly planar distribution of the 11 classical MW satellites has less
than a $0.5\%$ chance  to happen by chance if the
parent distribution is spherically symmetric, interpreting this as a
challenge to the Lambda Cold Dark Matter (LCDM) paradigm.
The same year it was recognized, using numerical simulations, that
this comparison was unfair given given that Dark Matter halos in LCDM
are expected to be triaxial and not spherical
Nevertheless, some estimates of the chances to find simulated satellites as
planar as in the MW 
were as low as $2\%$ \citep{2005ApJ...629..219Z} while others expected
planar satellite configurations in every DM halo
\citep{2005MNRAS.363..146L}.

@@

Trying to reconcile and understand the seemly conflicting or inconclusive
results on the MW has its origin on the frequentist fashion generally used to
compute probabilities.
Usually, the process starts by building a high level parent sample in the
simulation and then counting how many elements has a subset
meeting some criteria. 
The first inconvenient with this approach is that the probability
estimate is made against whatever turns out to be typical in each
simulation. 
Comparing results across simulations would require first
characterizing all the simulations at the high level parent samples,
something that is difficult to do in practice.
The second inconvenient is that it is hard to find systems that fully resemble
the LG, this means that the high level parent sample has a small size
making it hard to find elements that fit some criteria.

In this paper we control the first effect by setting as direct 
point of reference an spherical satellite distribution and not the
simulations themselves.
The spherical satellite distribution is built from the data itself
(observational or simulated) by randomizing its angular position
around the central galaxy and keeping its radial distance fixed
\citep{2017AN....338..854P}. 
We characterize each satellite distribution in terms of the
scalars describing its deviation from the spherical distribution.
We then build an explicit analytic
probability distribution of the deviations from asphericity; this
solves the problem of having a common reference point to compare
simulations. 
We then use different simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations, overcoming the problem of having a small
number of systems in the parent simulations.

@@

In Section \ref{sec:DataSamples} we list the sources of the observational and
simulated data to be used throughout the paper.
In Section \ref{sec:SpatialMeasurements} we describe the methods we use to build halo pairs and 
characterize its satellite distributions.
In Section \ref{sec:results} we present our results to finally
conclude in Section \ref{sec:conclusions}.

>>>>

This started with the suggestion of a Magellanic Plane, a flattened
structure of satellite galaxies and globular clusters around the MW,
first suggested by \cite{1976RGOB..182..241K} and \cite{1976MNRAS.174..695L}.  
\cite{2005A&A...431..517K} then quantified that the
highly planar distribution of the 11 classical MW satellites has less
than a $0.5\%$ chance  to happen by chance if the
parent distribution is spherically symmetric, interpreting this as a
challenge to the Lambda Cold Dark Matter (LCDM) paradigm.
The same year it was recognized, using numerical simulations, that
this comparison was unfair given that Dark Matter halos in LCDM
are expected to be triaxial and not spherical
Nevertheless, some estimates of the chances to find simulated satellites as
planar as in the MW 
were as low as $2\%$ \citep{2005ApJ...629..219Z} while others expected
planar satellite configurations in every DM halo
\citep{2005MNRAS.363..146L}.

@@

Some of the difficulty in trying to reconcile and understand the
seemly conflicting or inconclusive results on the MW has its origin on
the frequentist fashion generally used to compute probabilities.
Usually, this process starts by building a high level parent sample in the
simulation and then counting how many elements has a subset
meeting some criteria. This has two incovenients.
The first is that the probability estimate is made against whatever
turns out to be typical in each simulation. 
A fair comparison across simulations would require first
characterizing all the simulations at the high level parent samples,
something that is difficult to do in practice. 
The second inconvenient is that it is hard to find systems that fully resemble
the LG while having enough resolution to say something about the
satellites, this means that the high level parent sample has a small
size, making it hard to derive robust probabilities by counting.

In this paper we control the first effect by setting as a direct 
point of reference an spherical satellite distribution and not the
simulations themselves.
The spherical satellite distribution is built from the data itself
(observational or simulated) by randomizing the angular position
of each satellite around the central galaxy and keeping its radial
distance fixed \citep{2017AN....338..854P}. 
We characterize the satellites in terms of the scalars describing its
deviation from the spherical distribution. 
We then build an explicit analytic
probability distribution for the deviations from asphericity; this
solves the problem of having a common reference point to compare
simulations. 
We then use different simulations to estimate the parameters in this
distribution to later use it as a parent sample to generate any
desired number of samples that are by construction statistically
compatible with simulations; thus overcoming the problem of having a small
number of systems in the parent simulations.

@@

The rest of the paper is structured as follows. 
In Section \ref{sec:DataSamples} we list the sources of the observational and
simulated data to be used throughout the paper.
In Section \ref{sec:SpatialMeasurements} we describe the methods we use to build halo pairs and 
characterize its satellite distributions.
In Section \ref{sec:results} we present our results to finally
conclude in Section \ref{sec:conclusions}.
